Splitting of the homology of the punctured mapping class group

### 2020-05-01 arxiv version/Preliminaries.tex
Final version ready to appear in JTop

In this section we recollect some classical definitions and results about braid groups and mapping class groups.
% Let $\sg$ be a compact, smooth, orientable surface of genus $g$ with one parametrised
% boundary component $\partial \sg$. Let $\sgm$ be $\sg$ with a choice of $m$ distinct
% points in the interior: these points are called \emph{punctures} and are not assumed to be ordered.

@@

 The symmetric group $\mathfrak{S}_m$ acts freely on $F_m(\sg)$ by permuting the labels $1,\dots,n$ of a configuration;
 the orbit space
 \[
 F_m(\sg)/\mathfrak{S}_m
 \]
 is called the \emph{$m$-th unordered configuration space}
 of $\sg$ and is denoted by $C_m(\sg)$; this space is also a $2m$-dimensional orientable manifold
 (see figure \ref{fig:unordered}).

%  
%  We denote by $C_k(\sg)^{\infty}$ the one-point compactification of the space $C_k(\sg)$; the point
%  at infinity is denoted by $*$ and serves as basepoint.
\end{defn}

\begin{figure}\centering
\includegraphics[scale=0.8]{figures/unordered.pdf}
\caption{A configuration in the space $C_5(\Sigma_{1,1})$}
\label{fig:unordered}
\end{figure}

@@

 Similarly let $\Diff(\sgm;\partial\sgm)$ be the group of diffeomorphisms of $\sgm$ that fix $\partial(\sg)$
 pointwise and restrict to a permutation of the $m$ punctures: the \emph{mapping class group} of $\sgm$ is
 $\ggm=\pi_0(\Diff(\sgm;\partial\sgm))$.
 \end{defn}
 A classical result by Earle and Schatz \cite{EarleSchatz} ensures that the connected components
 of $\Diff(\sg)$ are contractible. In particular $B\Diff(\sg;\partial\sg)\simeq B\gg$ is
 a classifying space for
 the group $\gg$, i.e. an Eilenberg-MacLane space of type $K(\gg,1)$.

@@

In the whole article the genus $g\geq 0$ of the surfaces that we consider is supposed to be fixed,
% unless explicitly stated otherwise,
and we will abbreviate $\S=\sg$.
We denote by $\D$ the open disc $\mathring{\Sigma}_{0,1}$.

@@

\begin{figure}\centering
 \includegraphics[scale=0.8]{figures/SS'D.pdf}
 \caption{A splitting of $\S$ as $\S'\natural\bar\D$.}
\label{fig:SS'D}
\end{figure}

@@

\begin{defn}
\label{defn:muF}
For all $0\leq p\leq m$ there is a map
\[
 \mu\colon C_p(\D)\times C_{m-p}(\S')\to \cms
\]
which takes the union of configurations (see figure \ref{fig:defmu}):
 \[
  \mu\pa{\set{P_1,\dots,P_p};\set{P'_1,\dots,P'_{m-p}}}=\set{P_1,\dots,P_p,P'_1,\dots,P'_{m-p}}\in\cms.
 \]
\begin{figure}\centering
 \includegraphics[scale=0.7]{figures/defmu.pdf}
 \caption{The product of two configurations in $C_2(\D)$ and $C_4(\S')$.}
\label{fig:defmu}
\end{figure}

@@

Here $\epsilon\in H_0(C_1(\D))$ is the fundamental class, and for all
$k,m\geq 0$ we denote by $Q\colon H_k(C_m(\D))\to H_{2k+1}(C_{2m}(\D))$
the first Dyer-Lashof operation. In particular $Q^j\epsilon$ is the generator of
$H_{2^j-1}(C_{2^j}(\D))\simeq\Z_2$. See figure \ref{fig:monomial}

The isomorphism in equation \eqref{eq:Cohen} is
an isomorphism of bigraded rings. The left-hand side is a ring with the Pontryagin product,
% coming from the action of the little $2$-cubes,
and the right-hand side is a polynomial ring in infinitely many variables $\epsilon,Q\epsilon,Q^2\epsilon,\dots$.
The bigrading is given by the homological degree
$*$, that we call the \emph{degree},
and by the index $m$ of the connected component
on which the homology class is supported (informally, the number of points
involved in the construction of the homology class), that we call the \emph{weight}.

 \begin{figure}\centering
 \includegraphics[scale=0.7]{figures/monomial.pdf}
 \caption{The class in $H_5(C_9(\D))$ corresponding to the monomial $\epsilon\cdot(Q\epsilon)^2\cdot Q^2\epsilon$.}
\label{fig:monomial}
\end{figure}

@@

We consider in particular the one-point-compactification $\cms^{\infty}$ of $\cms$.

>>>>

In this section we recollect some classical definitions and results about braid groups and mapping class groups.

@@

 The symmetric group $\mathfrak{S}_m$ acts freely on $F_m(\sg)$ by permuting the labels $1,\dots,m$ of a configuration;
 the orbit space
 \[
 F_m(\sg)/\mathfrak{S}_m
 \]
 is called the \emph{$m$-th unordered configuration space}
 of $\sg$ and is denoted by $C_m(\sg)$; this space is also a $2m$-dimensional orientable manifold
 (see figure \ref{fig:unordered}).

\end{defn}

\begin{figure}\centering
\includegraphics[scale=0.8]{unordered.pdf}
\caption{A configuration in the space $C_5(\Sigma_{1,1})$}
\label{fig:unordered}
\end{figure}

@@

 Similarly let $\Diff(\sgm;\partial\sgm)$ be the group of diffeomorphisms of $\sgm$ that fix $\partial(\sg)$
 pointwise and restrict to a permutation of the $m$ punctures: the \emph{mapping class group} of $\sgm$ is
 $\ggm=\pi_0(\Diff(\sgm;\partial\sgm))$.
 \end{defn}
 A classical result by Earle and Schatz \cite{EarleSchatz} ensures that the connected components
 of $\Diff(\sg)$ are contractible: note that for $g=1$ we are using the fact
 that our surfaces have non-empty boundary. In particular $B\Diff(\sg;\partial\sg)\simeq B\gg$ is
 a classifying space for
 the group $\gg$, i.e. an Eilenberg-MacLane space of type $K(\gg,1)$.

@@

In the whole article the genus $g\geq 0$ of the surfaces that we consider is supposed to be fixed,
and we will abbreviate $\S=\sg$.
We denote by $\D$ the open disc $\mathring{\Sigma}_{0,1}$.

@@

\begin{figure}\centering
 \includegraphics[scale=0.8]{SS'D.pdf}
 \caption{A splitting of $\S$ as $\S'\natural\bar\D$.}
\label{fig:SS'D}
\end{figure}

@@

\begin{defn}
For all $0\leq p\leq m$ there is a map
\[
 \mu\colon C_p(\D)\times C_{m-p}(\S')\to \cms
\]
which takes the union of configurations (see figure \ref{fig:defmu}):
 \[
  \mu\pa{\set{P_1,\dots,P_p};\set{P'_1,\dots,P'_{m-p}}}=\set{P_1,\dots,P_p,P'_1,\dots,P'_{m-p}}\in\cms.
 \]
\begin{figure}\centering
 \includegraphics[scale=0.7]{defmu.pdf}
 \caption{The product of two configurations in $C_2(\D)$ and $C_4(\S')$.}
\label{fig:defmu}
\end{figure}

@@

Here $\epsilon\in H_0(C_1(\D))$ is the fundamental class, and for all
$k,m\geq 0$ we denote by $Q\colon H_k(C_m(\D))\to H_{2k+1}(C_{2m}(\D))$
the first Dyer-Lashof operation. In particular $Q^j\epsilon$ is the generator of
$H_{2^j-1}(C_{2^j}(\D))\simeq\Z_2$. See figure \ref{fig:monomial}.

The isomorphism in equation \eqref{eq:Cohen} is
an isomorphism of bigraded rings. The left-hand side is a ring with the Pontryagin product,
and the right-hand side is a polynomial ring in infinitely many variables $\epsilon,Q\epsilon,Q^2\epsilon,\dots$.
The bigrading is given by the homological degree
$*$, that we call the \emph{degree},
and by the index $m$ of the connected component
on which the homology class is supported (informally, the number of points
involved in the construction of the homology class), that we call the \emph{weight}.

 \begin{figure}\centering
 \includegraphics[scale=0.7]{monomial.pdf}
 \caption{The class in $H_5(C_9(\D))$ corresponding to the monomial $\epsilon\cdot(Q\epsilon)^2\cdot Q^2\epsilon$.}
\label{fig:monomial}
\end{figure}

@@

We will consider in particular the one-point-compactification $\cms^{\infty}$ of $\cms$.

### 2020-05-01 arxiv version/Ssequence.tex
Final version ready to appear in JTop

\section{Proof of Theorem \ref{thm:main}}
\ref{sec:sseq}
We will prove Theorem \ref{thm:main} by induction on $m$.
The case $m=0$ is trivial.

@@

In particular if we prove the vanishing of all differentials $\partial_r$
% exiting from the summand
% $H_k\pa{\gg;H_{p-l}(C_p(\D))\otimes H_{m-p}(C_{m-p}(\S'))}$
in the first spectral sequence, then also all differentials
$\partial_r$ exiting from this direct summand in the
second spectral sequence $E(m)$ must vanish.

% We observe that there is an isomorphism
% \[
%  H_k\pa{\gg;H_{p-l}(C_p(\D))\otimes H_{m-p}(C_{m-p}(\S'))}\simeq H_{p-l}(C_p(\D))\otimes H_k\pa{\gg; H_{m-p}(C_{m-p}(\S'))};
% \]
The differentials in the spectral sequence $H_*(C_p(\D))\otimes E(m-p)$
are obtained by tensoring the identity of $H_*(C_p(\D))$ with the differentials
of the spectral sequence $E(m-p)$; as $p\geq l\geq 1$ we know by inductive hypothesis
that the latter vanish. Theorem \ref{thm:main} is proved.

One can generalise Theorem \ref{thm:main} to orientable or non-orientable
surfaces with non-empty boundary, following the generalisation
of Theorem \ref{thm:Hbms*as*ggrep} discussed at the end of section \ref{sec:Actiongg}.

>>>>

\section{Proof of Theorem \ref{thm:main}}
We will prove Theorem \ref{thm:main} by induction on $m$.
The case $m=0$ is trivial.

@@

In particular if we prove the vanishing of all differentials $\partial_r$
in the first spectral sequence, then also all differentials
$\partial_r$ exiting from this direct summand in the
second spectral sequence $E(m)$ must vanish.

The differentials in the spectral sequence $H_*(C_p(\D))\otimes E(m-p)$
are obtained by tensoring the identity of $H_*(C_p(\D))$ with the differentials
of the spectral sequence $E(m-p)$; as $p\geq l\geq 1$ we know by inductive hypothesis
that the latter vanish. Theorem \ref{thm:main} is proved.

One can generalise Theorem \ref{thm:main} to orientable or non-orientable
surfaces with non-empty boundary, following the generalisation
of Theorem \ref{thm:Hbms*as*ggrep} discussed at the end of Section \ref{sec:Actiongg}.
